Identifying covariate traits of trance and possession phenomena in the Ethnographic

### 2022-10-20 Readme.Rmd
revamped analysis

Here is the distribution of the various descriptions for trance and possession in our training sample. I gave the descriptions shorter names as well, to make things easier later on:

```{r }
sccs %>% 
  filter(var_id == 'EA112') %>% 
  count(var_name,var_description) %>% 
  rename(name = var_name, description = var_description, count = n) %>% 
  add_row(name = 'Total', description = '', count = 151) %>% 
  kable('simple')
```

We see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form.

We compare this with how our model categorises each society. This is our model's confusion matrix. The rows are the actual descriptions, the columns are the predicted descriptions. So a good model would have large counts on the diagonal (where it got the label right) and low values elsewhere (where it didn't). The last two columns give the model's error as a probability (0-1) and as a rate (as n/m).

```{r }
cm1a = read_tsv('models/conf_matrix11model.tsv')
cm1a %>% 
  mutate(original = c(names(cm1a)[!names(cm1a) %in% c('Error','Rate')],'Total')) %>% 
  relocate(original) %>% 
  kable('simple', digits = 2)
```

These results are not very impressive. The model gets some of the larger categories less wrong but it is completely off the mark on the smaller ones. (It gets only one society right in the "no trance or possession" category and all of them wrong in the "possession-only" category!)

Focussing on the relationship between category sizes and model accuracy, we can tell two stories here. The first one is that the model is terrible and it is only likely to do better with the larger categories because of their size (so that the baseline probability of category membership is higher). The second story is largely the same but with some hopeful notes.

We can say that the way trance and possession phenomena are defined in the Atlas is very fragmentary. It involves at least two external sources of noise. First, for a category like "trance only, no possession", maybe the curators of the Atlas put very disparate psychosocial phenomena into this category, which haven't a lot to do with each other at all. In this case, it is no small wonder that we can't capture this category algorithmically. Maybe it's all arbitrary!

Second, maybe, even for the larger categories, the boundaries are porous. Maybe it's not entirely clear whether, in a given society, only possession explains trance phenomena. Maybe the fieldworkers missed something! Maybe they disagree on their definitions of things (e.g. some say that a trance phenomenon in a given society is not caused by possession ergo it is not trance at all but something else).

We can work around some of these issues by re-defining our outcome variable. We split our descriptions into two categories, based on whether they capture possession trance or not. This would look like this:

```{r }
d %>% 
  filter(var_title == 'Trance states') %>% 
  distinct(var_description) %>% 
  rename(description = var_description) %>% 
  add_column(`this is...` = c('something else','something else','possession trance','possession trance','possession trance','something else','something else','possession trance')) %>% 
  kable('simple')
```

Now we have two categories that are roughly the same size and are perhaps less conceptually ambiguous. We refit the model, but this time it only tries to guess whether a society has possession trance or something else (including no trance or possession at all).

## 6. Results on the training sample, part II

We look at the model's confusion matrix:

```{r }
cm4a = read_tsv('models/conf_matrix41model.tsv')
cm4a %>% 
  mutate(original = c(names(cm4a)[!names(cm4a) %in% c('Error','Rate')],'Total')) %>% 
  relocate(original) %>% 
  kable('simple', digits = 2)
```

This is a lot better. Granted, there are only two categories, but the error rate is much lower. This means our model can more or less use the societal covariants to guess whether a society, by the Atlas' own terms, will have possession trance phenomena (1) or not (0).

Having established this, we can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available, of course) -- including the SCCS subset. Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.

Here is the confusion matrix for all `r length(unique(d$soc_id))` societies in the Atlas where trance and possession information are available:

```{r }
cm4b = read_tsv('models/conf_matrix41test.tsv')

cm4b %>% 
  add_column(original = c(0,0,0,1,1,0,1,1)) %>% 
  group_by(original) %>% 
  summarise(`0` = sum(`0`), `1` = sum(`1`)) %>% 
  kable('simple')

# d %>% 
#   filter(var_id == 'EA112') %>% 
#   mutate(
#     possession_trance_present = case_when(
#       code %in% c(3,4,6,7) ~ T,
#       code %in% c(1,2,5,8) ~ F
#         )
#   ) %>% 
#   count(possession_trance_present) %>% 
#   kable('simple')
```

This is much worse than the training data only but that is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view. And it is still not terrible! It gets 468/`r length(unique(d$soc_id))` societies right. There are 281 versus 377 societies in the data with and without possession trance. If it assumed that no societies had possession trance, it would get 377 data right, and 468 is considerably larger number than that. This means the model is doing something right.

Now that we established that the model does find some purchase, we take a look at the variables that were most important in predicting the presence of possession trance in a given society. I will only show societal covariants with a high enough importance:

```{r }
varimp4 = read_tsv('models/varimp41.tsv')
vars2 = varimp4 %>% 
  filter(relative_importance > 15) %>%
  mutate(ID = str_extract(variable, '^EA...')) %>% 
  left_join(vars, by = "ID") %>% 
  select(ID, Name, relative_importance)

@@

What do we see? The five most important driving factors are the prevalence of slavery, family complexity, the complexity of class structure, premarital norms for women, and settlement patterns. Various aspects of social complexity and filial and consanguineal organisation also make the list. We can cross-reference this with our first table, with pointers in the literature. We add a column indicating whether the Atlas seems to support each specific claim. (This we do in a very haphazard manner!)

| Relationship|Reference|Our model|
|:-|:-|:-|
|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|not directly, but more complexity means possession trance is more likely|
|women's social participation|@bourguignon1973religion|not directly, but premarital sexual norms can be a proxy|
|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
|main mode of subsistence|@bourgouignon1977|not really|
|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but family and kinship seems to do something|
|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
|monotheism|@wood2018complexity|absolutely, the variable is right there|

At this point, we can say, okay, the model says that these various societal covariates are meaningful in the Atlas. But the direction of the individual relationships is not clear. If our analysis finds that increased prevalence of slavery makes possession trance less likely, it's clearly at odds with the existing literature and probably doesn't make sense.

## 7. Making sense: a figure

@@

```{r fig.width=10,fig.height=15}
d_out = d %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    possession_trance_present = case_when(
      code %in% c(3,4,6,7) ~ 'present',
      code %in% c(1,2,5,8) ~ 'absent'
        )
      ) %>% 
  select(soc_id,possession_trance_present)

d_in = d %>%
  filter(var_title %in% vars2$Name, !is.na(code)) %>% 
  mutate(
    predictor = fct_reorder(var_name, -as.double(recoded_value)),
    var_title = fct_relevel(var_title, vars2$Name)
    ) %>% 
  select(soc_id,var_definition,var_title,predictor)

@@

For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up. Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent. Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements.

The societal variables that do not, strictly speaking, measure social organisation are less obvious. Apparently cousin marriage is good for possession trance, and so is having a lot of words for your various cousins. (We might suspect the latter is here because it is heavily correlated with the former.) Active, moral high gods are bad for possession trance. Prohibitions on pre-married sexual behaviour for women go with more prevalent possession trance, but this pattern is rather variable.

## 8. Summary

>>>>

Here is the distribution of the various descriptions for trance and possession in our training sample.

```{r }
trance = sccs %>% 
  filter(var_id == 'EA112') %>% 
  count(var_name,code,var_description) %>% 
  rename(name = var_name, description = var_description, count = n)

trance %>% 
  select(-code) %>% 
  add_row(name = 'Total', description = '', count = 151) %>% 
  kable('simple')
```

We see that only about 10% of societies have no trance or possession phenomena. From the rest, a majority have possession trance in some form. Following @bourguignon1968 we can reduce this to a binary: 81 societies have possession trance, 70 have other forms of trance and possessior or show neither.

```{r }
trance %>% 
  mutate(
    possession_trance = case_when(
      code %in% c(3,4,6,7) ~ 'possession trance',
      code %in% c(1,2,8,5) ~ 'other'
    ) %>% 
      fct_relevel('possession trance')
  ) %>% 
  group_by(possession_trance) %>% 
  summarise(count = sum(count)) %>% 
  kable('simple')
```

We compare this with how our best model categorises each society. This is our model's confusion matrix. `0` is absence of possession trance, `1` is presence of possession trance. The rows add up to values in the original, the columns are the model's predictions. The last two columns show the error rate. The model's accuracy on absence of possession trance is relatively low, though higher than chance (54 out of 80, p < 0.01 using a binomial test with a true p of .529). It is mostly accurate capturing the presence of possession trance.

```{r }
cm1a = read_tsv('models/conf_matrix11model.tsv')
cm1a %>% 
  mutate(original = c(names(cm1a)[!names(cm1a) %in% c('Error','Rate')],'Total')) %>% 
  relocate(original) %>% 
  kable('simple', digits = 2)

# precision = true 1/false 1
# recall = true 1/true1 + false0
precision = 68 / 94
recall = 68 / 71
f = (2 * precision * recall) / (precision + recall)
```

The model has an F-measure of `r round(f,2)`, with a penalty on overgeneralising possession trance. Though comparisons are hard to make, this is not particularly accurate for a binary categorisation problem. We did expect this, though, as it is unlikely that any complex cultural phenomenon could be reduced to a small number of cross-cultural predictors.

### 6. Results on the EA data

We can ask the model to make predictions for the entire Ethnographic Atlas (where trance and possession data are available) -- including the SCCS subset. Generally, you want to separate the training and test sets in Machine Learning and not do this, but the 'everything except the SCCS' subset of the Atlas would be an odd test set.

Here is the confusion matrix for all `r length(unique(ld$soc_id))` societies in the Atlas where trance and possession information are available:

```{r }
cm1b = read_tsv('models/conf_matrix11test.tsv')

cm1b %>% 
  kable('simple')

# precision = true 1/false 1
# recall = true 1/true1 + false0
precision2 = 221 / 389
recall2 = 221 / 281
f2 = (2 * precision2 * recall2) / (precision2 + recall2)

```

The F-measue of this model is `r round(f2,2)`. Accuracy here is much worse than for the training data only. That is at least partly because the model is ignorant of the phylogenetic signal which might result in patterns that are unexpected from a purely correlational point of view. However, the model is still more accurate than it would be by chance.

We take a look at the variables that were most important in the model in predicting the presence of possession trance in a given society, with a cutoff of 10/47.

```{r }
varimp1 = read_tsv('models/varimp11.tsv')
vars2 = varimp1 %>% 
  slice(1:10) %>% 
  mutate(ID = str_extract(variable, '^EA...')) %>% 
  left_join(vars, by = "ID") %>% 
  select(ID, Name, relative_importance)

@@

We now look at the direction of these effects (does e.g.\ increased jurisdictional hierarchy go with more or less prevalent possession trance?).

@@

For the most important variables, we see a clear correllation between the presence of possession trance and the variable scaling up. Societies in which slavery is more prevalent have more possession trance than those where it is less prevalent. Same is true for societies with more complex family structures, more social complexity, more complex settlement patterns, and larger settlements. The societal variables that do not, strictly speaking, measure social organisation are less obvious.

```{r fig.width=10,fig.height=15}
d_out = ld %>% 
  filter(var_id == 'EA112') %>% 
  mutate(
    possession_trance_present = case_when(
      code %in% c(3,4,6,7) ~ 'present',
      code %in% c(1,2,5,8) ~ 'absent'
        )
      ) %>% 
  select(soc_id,possession_trance_present)

d_in = ld %>%
  filter(var_title %in% vars2$Name, !is.na(code)) %>% 
  mutate(
    predictor = fct_reorder(var_name, -as.double(recoded_value)),
    var_title = fct_relevel(var_title, vars2$Name)
    ) %>% 
  select(soc_id,var_definition,var_title,predictor)

@@

## 7. Summary

These results go some way in confirming hypotheses on the cross-cultural correlates of possession trance in the literature. @bourguignon1968 explicitly tests the variables of slavery, jurisdictional hierarchy, settlement patterns, and kin groups. @shaara1992 point to a high value placed on female virginity at marriage, captured here by norms of premarital sexual behaviour for girls. These are all important variables in our model. @greenbaum1973 and @shaara1992 point to sexual division of labour, while @wood2018complexity to the role of moralising high gods, neither of which is relevant in our model.

This brings us back to the original predictions in the table above.

| Relationship|Reference|Our model|
|:-|:-|:-|
|less complexity$\rightarrow{}$trance w/o possession|@bourguignon1968|the same measures of complexity are relevant|
|women's social participation|@bourguignon1973religion|labour divisions are not relevant|
|increased role differentiation and slavery practices$\rightarrow$trance and possession|@greenbaum1973|slavery, social and class complexity are relevant|
|main mode of subsistence|@bourgouignon1977|not relevant|
|increased social participation$\rightarrow{}$possession|@swanson1978|indirectly, through domestic and social complexity|
|more complexity$\rightarrow{}$trance w/ possession|@winkelman1986|yes, see above|
|cognatic kinship$\rightarrow{}$more possession and trance|@shaara1992|not directly, but kinship structure is relevant|
|decreased social participation$\rightarrow{}$possession|@douglas2004natural|yes, see above|
|monotheism|@wood2018complexity|not relevant in our model|
